Variance reduction for MCMC methods via martingale representations

### 2019-12-12 aap/Revision/vr-mcmc-revision.tex
Revision

\begin{description}
\item[(H1)][\textbf{Lipschitz continuity}] Assume that $\nabla U$ is Lipschitz, that is, there exists $L\geq 0$ such that 
\begin{eqnarray*}
\|\nabla U(x)-\nabla U(y)\|\leq L\|x-y\|, \quad x,y\in \mathbb{R}^d.
\end{eqnarray*}

\item[(H2)][\textbf{Convexity outside a ball}] There exist $K>0,$ $M>0$ and $m>0$ such that for any $x\not\in B_K(0)	$ and any $y\in \mathbb{R}^d$ it holds 

@@

\langle D^2U(x)y,y\rangle \geq m\|y\|^2, \quad \|D^3 U(y)\| \leq M. 

@@

from $(\mathbb R^d,\mathcal B(\mathbb R^d))$
into $(\mathbb R^d,\mathcal B(\mathbb R^d))$

@@

which, finally, provides the (random) measure
$Q_\gamma^{N+1,p}(X_N,\cdot)$

@@

As for the estimator~\eqref{eq:29032018a3}, we introduce lemma of conditional variance decomposition.
It follows from~\eqref{eq:29032018a5}, Theorem~\ref{prop:29032018a1} and independence condition for the sequence $(Z_p)_{p \geq 1}$.

@@

\mathsf{Var}\left[\left.\pi_{K,n}^{N}(f)\right|\mathcal G_N\right]

@@

\mathsf{E}\left[\left.\left|\bar a_{l,\mathbf{k}}(X_{l-1})\right|^{2}
\right|\mathcal G_N\right].

@@

@@

for all \(\mathbf{k}\) with \(0<\|\mathbf{k}\|\leq  2.\) Moreover assume that there exist $K>0,$ $M>0$ and $m>0$ such that for any $x\not\in B_K(0)$ and any $y\in \mathbb{R}^d$ it holds
\langle \nabla \mu(x)y,y\rangle \geq m\|y\|^2.
Let $(\gamma_k)_{k\in\mathbb N}$ be a sequence of positive

@@

\mathsf{Var}\left[\left.\pi_{K,n}^{N}(f)\right|\mathcal G_N\right]\lesssim \frac{1}{\Gamma^2_{N+2,N+n+1}},

@@

\Bigg|\mathcal G_N \right\}.

@@

\Bigg|\mathcal G_N\right].

@@

In what follows, we use the notation
\begin{equation}
\label{eq:definition-alpha}
\alpha_k=1-\gamma_k b_\mu,\quad k\in\mathbb N.
\end{equation}

\begin{remark}\label{rem:10062018a1}
Notice that, under~\eqref{eq:17042018a1},
not only~\eqref{eq:06062018a3} but also
\begin{equation}\label{eq:10062018a1}
\sum_{r=j}^\infty\gamma_r \prod_{k=j}^r \alpha_k\le\ol C_2
\end{equation}
is satisfied with the same constant~$\ol C_2$
(which is $C$ of~\eqref{eq:17042018a1})
simultaneously for all $j\in\mathbb N$.
Below this will allow us to apply
Lemma~\ref{lem:06062018a1}
to bound double indexed sequences
$(x_{j,p})_{j\ge1,p\ge j}$
satisfying
$$
0\le x_{j,p}\le\alpha_p x_{j,p-1}+\gamma_p \epsilon_{j,p},\quad p\ge j+1,
$$
with suitable $(\epsilon_{j,p})_{j\ge1,p\ge j+1}$
and the constant $\ol C_2$ in~\eqref{eq:10062018a1}
being independent of~$j$.
\end{remark}

\begin{lem}\label{lem:06062018a2}
Under assumptions of Theorem~\ref{th:mr},
for all natural $r\le K$ and $l\le p$, there exist constants $C_r$ (not depending on $l$ and $p$) such that
\begin{equation}
\label{eq:08062018b2}
\left|\partial_{y_l}^r X_p \right| \le C_r\prod_{k=l+1}^p \alpha_k \,,
\end{equation}
where $\partial_{y_l}^r X_p$ is defined in \eqref{eq:definition-differential-f-p}. Moreover, we can choose $C_1=1$.
\end{lem}
\begin{proof}
The proof is along the same lines as Lemma~\ref{lem:06062018a3}.
\end{proof}
\begin{lem}\label{lem:06062018a3}
Under assumptions of Theorem~\ref{th:mr},
for all natural $r\le K$, $j\ge l$ and $p>j$, we have
\begin{equation}\label{eq:08062018b3}
\left|\partial_{y_j} \partial_{y_l}^r X_p\right|
\le c_r\prod_{k=l+1}^p \alpha_k
\end{equation}
with some constants $c_r$
not depending on $j$, $l$ and $p$,
while, for $p\le j$, it holds
$\partial_{y_{j}}\partial_{y_{l}}^{r}X_{p}=0$.
\end{lem}

@@

with a suitable constant
(we can take, e.g., $\const=\frac{B_\mu}{(1-\gamma_1 b_\mu)^2}$).

@@

with some constant ``$\const$'',
which is, in fact, $\frac1{(1-\gamma_1 b_\mu)^2}$ times
the expression involving

@@

\left(\sum_{p=j}^{q}\gamma_{p+1}\prod_{k=l+1}^{p}\alpha_{k}\right)^{2}

@@

\frac{A_{K}^{2}}{(1-\gamma_1 b_\mu)^2}

@@

\frac{A_{K}^{2}}{(1-\gamma_1 b_\mu)^3}

@@

\frac{A_{K}^{2}}{(1-\gamma_1 b_\mu)^3}C^3=B_K,

>>>>

\begin{description}
\item[(H1)][\textbf{Lipschitz continuity}] Assume that $\nabla U$ is Lipschitz, that is, there exists $L_U\geq 0$ such that 

@@

|\nabla U(x)-\nabla U(y)|\leq L_U|x-y|, \quad x,y\in \mathbb{R}^d.
\item[(H2)][\textbf{Convexity outside a ball}] There exist $K_U>0,$ $M_U>0$ and $m_U>0$ such that for any $x\not\in B_{K_U}(0)	$ and any $y\in \mathbb{R}^d$ it holds 
\langle D^2U(x)y,y\rangle \geq m_U\|y\|^2, \quad \|D^3 U(y)\| \leq M_U. 

@@

@@

which, finally, provides the  measure
$Q_\gamma^{p}(x_0,\cdot)=Q_\gamma^{p,1}$

@@

As for the estimator~\eqref{eq:29032018a3}, 
it follows from~\eqref{eq:29032018a5}, Theorem~\ref{prop:29032018a1} and independence condition for the sequence $(Z_p)_{p \geq 1}$.

@@

\mathsf{Var}\left[\pi_{K,n}^{N}(f)\right]

@@

\mathsf{E}\left[\left|\bar a_{l,\mathbf{k}}(X_{l-1})\right|^{2}\right].

@@

Assume (H1) and (H2).

@@

for all \(\mathbf{k}\) with \(0<\|\mathbf{k}\|\leq  2.\) Let $(\gamma_k)_{k\in\mathbb N}$ be a sequence of positive

@@

\mathsf{Var}\left[\pi_{K,n}^{N}(f)\right]\lesssim \frac{1}{\Gamma^2_{N+2,N+n+1}},

@@

\right\}.

@@

\Big|X_{l-1}\right)\right].

@@

In what follows, we use the notation
\begin{equation}
\label{eq:definition-alpha}
\alpha_k=1-\gamma_k \mu'(X_{k-1}),\quad k\in\mathbb N.
\end{equation}

\begin{remark}\label{rem:10062018a1}

Below this will allow us to apply
Lemma~\ref{lem:06062018a1}
to bound double indexed sequences
$(x_{j,p})_{j\ge1,p\ge j}$
satisfying
$$
0\le x_{j,p}\le\alpha_p x_{j,p-1}+\gamma_p \epsilon_{j,p},\quad p\ge j+1,
$$
with suitable $(\epsilon_{j,p})_{j\ge1,p\ge j+1}$.
\end{remark}
The assumption (H1) implies that  $|\mu'(x)|\leq B_\mu$ for some constant $B_\mu>0$ and all $x\in \mathbb{R}^d$. Without loss of generality we suppose that $\gamma_N B_\mu<1.$
\begin{lem}\label{lem:06062018a2}
Under assumptions of Theorem~\ref{th:mr},
for all natural $r\le K$ and $l\le p$, there exist constants $C_r$ (not depending on $l$ and $p$) such that
\begin{equation}
\label{eq:08062018b2}
\left|\partial_{y_l}^r X_p \right| \le C_r\prod_{k=l+1}^p (1-\gamma_k\mu'(X_{k-1}))\quad \text{a.s.} 
\end{equation}
where $\partial_{y_l}^r X_p$ is defined in \eqref{eq:definition-differential-f-p}. Moreover, we can choose $C_1=1$.
\end{lem}
\begin{lem}\label{lem:06062018a3}
Under assumptions of Theorem~\ref{th:mr},
for all natural $r\le K$, $j\ge l$ and $p>j$, we have
\begin{equation}\label{eq:08062018b3}
\left|\partial_{y_j} \partial_{y_l}^r X_p\right|
\le c_r\prod_{k=l+1}^p (1-\gamma_k\mu'(X_{k-1})), \quad \text{a.s.}
\end{equation}
with some constants $c_r$
not depending on $j$, $l$ and $p$,
while, for $p\le j$, it holds
$\partial_{y_{j}}\partial_{y_{l}}^{r}X_{p}=0$.
\end{lem}

@@

with a suitable constant.

@@

with some constant ``$\const$'' depending on

@@

\mathsf{E}\left(\sum_{p=j}^{q}\gamma_{p+1}\prod_{k=l+1}^{p}\alpha_{k}\right)^{2}

@@

\frac{A_{K}^{2}}{(1-\gamma_1 B_\mu)^2}

@@

\frac{A_{K}^{2}}{(1-\gamma_1 B_\mu)^3}

@@

\frac{A_{K}^{2}}{(1-\gamma_1 B_\mu)^3}C^3=B_K,
